---
title: "Injury Risk in New York City Traffic Crashes"
subtitle: "Time, location, and reported crash outcomes, 2013-2025"
author: "Hao Yang"
date: last-modified
---

```{r}
#| label: setup
#| include: false

table_dir <- here::here("outputs", "tables")
figure_dir <- here::here("outputs", "figures")

read_result <- function(name) {
  readr::read_csv(
    file.path(table_dir, name),
    show_col_types = FALSE
  )
}

annual <- read_result("annual_summary.csv")
borough <- read_result("borough_summary.csv")
hourly <- read_result("hourly_summary.csv")
rush <- read_result("rush_summary.csv")
sample_balance <- read_result("sample_balance.csv")
bootstrap <- read_result("bootstrap_summary.csv")
odds_ratios <- read_result("odds_ratios.csv")
model_fit <- read_result("model_fit.csv")

full_risk <- sample_balance |>
  dplyr::filter(.data$sample == "Full data")
rush_risk_difference <- bootstrap |>
  dplyr::filter(.data$metric == "rush_injury_rate_difference")
```

## Executive summary

This project asks how crash timing and location are associated with the
probability that a police-reported New York City crash results in at least one
injury. It uses 2.13 million crash records from complete calendar years
2013-2025 and a reproducible 5% sample within each year for computationally
intensive inference.

Across the analysis period, `r scales::percent(full_risk$injury_rate, accuracy = 0.01)`
of reported crashes involved at least one injury. The year-stratified bootstrap
estimate for the rush-hour minus non-rush-hour injury rate difference is
`r scales::percent(rush_risk_difference$estimate, accuracy = 0.01)`, with a
percentile interval from `r scales::percent(rush_risk_difference$conf_low, accuracy = 0.01)`
to `r scales::percent(rush_risk_difference$conf_high, accuracy = 0.01)`.

These are associations within reported crash records. They are not causal
effects, and they do not measure risk per trip or mile traveled.

## Research question

> How are time of day and borough associated with the probability that a
> police-reported NYC crash results in at least one injury?

The analysis focuses on one primary binary outcome, `injury_crash`, and uses
mean injuries per crash as a secondary severity measure. This keeps the
descriptive summaries, bootstrap estimands, and logistic regression aligned.

## Data and design

The source is NYC Open Data's
[Motor Vehicle Collisions - Crashes](https://data.cityofnewyork.us/Public-Safety/Motor-Vehicle-Collisions-Crashes/h9gi-nx95)
table. Each row represents one police-reported crash. The pipeline retrieves
only variables used in the analysis and caches the official data by calendar
year.

The cleaning rules are explicit:

1. Deduplicate records by collision ID.
2. Parse dates and clock times without relying on implicit type conversion.
3. Retain only complete calendar years 2013-2025.
4. Preserve missing injury counts instead of silently treating them as zero.
5. Restrict coordinates to a prespecified NYC bounding box for spatial plots.
6. Define rush hour before analysis as 7:00-9:59 and 16:00-18:59.

The full data are used for descriptive summaries. Logistic regression and
bootstrap inference use a 5% random sample within each year with seed 2026.
Bootstrap resampling is also stratified by year, preserving the observed annual
sample sizes.

```{r}
#| label: tbl-sample-balance
#| echo: false

sample_balance |>
  dplyr::mutate(
    injury_rate = scales::percent(.data$injury_rate, accuracy = 0.01),
    borough_recorded = scales::percent(.data$borough_recorded, accuracy = 0.01),
    coordinates_valid = scales::percent(.data$coordinates_valid, accuracy = 0.01),
    mean_injuries = round(.data$mean_injuries, 4)
  ) |>
  dplyr::rename(
    Sample = sample,
    Crashes = crashes,
    `Mean injuries` = mean_injuries,
    `Injury rate` = injury_rate,
    `Borough recorded` = borough_recorded,
    `Valid coordinates` = coordinates_valid
  ) |>
  knitr::kable(
    caption = "Full-data and year-stratified sample comparison",
    format.args = list(big.mark = ",")
  )
```

Only `r sum(annual$missing_injury_count)` records in the analysis window have a
missing total injury count. Although recoding them to zero would have little
numerical effect here, preserving them as missing avoids an unsupported data
assumption.

## Annual pattern

```{r}
#| label: fig-annual
#| fig-cap: "Annual share of reported crashes with at least one injury."
#| echo: false

knitr::include_graphics(
  file.path(figure_dir, "annual_injury_rate.png")
)
```

The recorded injury share changes sharply across the study window, especially
after 2020. That pattern could reflect changes in traffic conditions, crash mix,
reporting, or data-generating processes. Without exposure data and a research
design that isolates those mechanisms, the annual series should not be treated
as the effect of a particular policy or event.

## Time of day

```{r}
#| label: fig-hourly
#| fig-cap: "Injury rate by crash hour. Shading marks the prespecified rush-hour windows."
#| echo: false

knitr::include_graphics(
  file.path(figure_dir, "hourly_injury_rate.png")
)
```

The rush-hour comparison is reported as a difference, with rush hour first.
A negative value means the observed injury rate or mean injury count is lower
during the prespecified rush windows.

```{r}
#| label: tbl-bootstrap
#| echo: false

metric_labels <- c(
  mean_injuries = "Mean injuries per crash",
  injury_rate = "Share with at least one injury",
  rush_mean_injury_difference = "Rush minus non-rush mean injuries",
  rush_injury_rate_difference = "Rush minus non-rush injury rate"
)

bootstrap |>
  dplyr::filter(.data$metric %in% names(metric_labels)) |>
  dplyr::mutate(
    metric = dplyr::recode(.data$metric, !!!metric_labels),
    dplyr::across(
      c(estimate, bootstrap_se, conf_low, conf_high),
      ~ round(.x, 5)
    )
  ) |>
  dplyr::select(
    Estimand = metric,
    Estimate = estimate,
    `Bootstrap SE` = bootstrap_se,
    `95% CI lower` = conf_low,
    `95% CI upper` = conf_high
  ) |>
  knitr::kable(
    caption = "Year-stratified bootstrap estimates"
  )
```

Traffic speed is one possible explanation for a rush-hour difference, but this
dataset does not measure speed or congestion. The result therefore remains an
unadjusted descriptive association.

## Borough pattern

```{r}
#| label: fig-borough
#| fig-cap: "Injury rate by borough with year-stratified bootstrap intervals."
#| echo: false

knitr::include_graphics(
  file.path(figure_dir, "borough_injury_rate.png")
)
```

Borough comparisons are conditional on crashes with a recorded borough. They
do not adjust for traffic volume, road design, travel mode, speed, or the mix of
reported crashes. The project therefore avoids attributing the observed
Manhattan difference to Vision Zero or enforcement without additional evidence.

## Adjusted model

A logistic regression models whether a crash produced at least one injury.
Predictors are time band, borough, weekend status, and calendar-year fixed
effects. Midday, Manhattan, weekdays, and 2013 are reference categories.

```{r}
#| label: fig-model
#| fig-cap: "Adjusted odds ratios for time, borough, and weekend indicators. Calendar-year fixed effects are not displayed."
#| echo: false

knitr::include_graphics(
  file.path(figure_dir, "model_odds_ratios.png")
)
```

```{r}
#| label: tbl-model-fit
#| echo: false

model_fit |>
  dplyr::mutate(
    observations = scales::comma(.data$observations),
    dplyr::across(c(aic, null_deviance, residual_deviance), ~ round(.x, 1))
  ) |>
  dplyr::rename(
    Observations = observations,
    AIC = aic,
    `Null deviance` = null_deviance,
    `Residual deviance` = residual_deviance
  ) |>
  knitr::kable(caption = "Logistic regression fit summary")
```

The coefficients are adjusted associations, not causal effects. Calendar year
is treated as a categorical fixed effect so the model does not impose a single
linear trend over a period with a visibly nonlinear outcome pattern. In-sample
agreement between fitted and observed borough averages is not presented as
evidence of out-of-sample calibration.

## Spatial distribution

```{r}
#| label: fig-map
#| fig-cap: "Two-dimensional bins summarize crash density without plotting nearly 100,000 overlapping points."
#| echo: false

knitr::include_graphics(
  file.path(figure_dir, "crash_density.png")
)
```

The density map is descriptive. It shows where reported crashes are
concentrated, but it cannot identify high-risk locations without a denominator
such as trips, vehicle miles, pedestrian counts, or road mileage.

## Limitations

1. The data include police-reported crashes and may not represent unreported
   events.
2. Exposure is unavailable, so the analysis cannot estimate risk per trip,
   mile driven, pedestrian, cyclist, or resident.
3. Borough is missing for a substantial share of records, and missingness may
   not be random.
4. Injury counts, locations, and contributing factors may contain reporting or
   measurement error.
5. Traffic volume, weather, road design, speed, and vehicle mix are unavailable
   or not modeled and may confound the reported associations.
6. A very large sample produces narrow intervals, but precision does not resolve
   selection, measurement, or causal-identification problems.

## Reproducibility

The repository contains the official API query, local-CSV import route,
cleaning functions, fixed analysis window, random seed, bootstrap code, model
specification, unit tests, and scripts needed to rebuild every displayed result.
Crash-level files are excluded from version control because they are large and
can be reacquired from the official source.
